Notes on Validated Model Counting \\ Version of \today

### 2022-04-25 doc/mc-notes.tex
Ready for discussion

A {\em Boolean function} $f:2^X \rightarrow \{0,1\}$ can be
characterized by the set of assignments for which the function
evaluates to 1: $\interpset{f} = \{ \interp | f(\interp) = 1\}$.  Let
$\one$ denote the Boolean function that assigns value 1 to every
assignment, and $\zero$ denote the assignment that assigns value 0 to
every assignment.  These are characterized by the universal and empty
assignment sets, respectively.

From
this we can define the {\em complement} of function $f$ as the function
$\boolnot f$ such that
$\interpset{\boolnot f} = \{ \interp | f(\interp) = 0\}$.
We can also define the conjunction and disjunction operations over functions $f_1$ and $f_2$ as characterized by the sets
$\interpset{f_1 \booland f_2} = \interpset{f_1} \cap \interpset{f_2}$ and
$\interpset{f_1 \boolor f_2} = \interpset{f_1} \cup \interpset{f_2}$.

For assignment $\interp$ and a Boolean formula $\phi$ over $X$, we
use the notation $\interp\subs{x_i}{\phi}$ to denote the assignment
$\interp'$, such that $\interp'(x_j) = \interp(x_j)$ for all $j \not = i$
and $\interp'(x_i) = \interp(\phi)$, where $\interp(\phi)$ is the value obtained by evaluating formula $E$ with each variable assigned the value given by $\interp$.
In particular, the notation
$\interp(\subs{x_i}{\obar{x}_i})$
indicates the assignment in which the value
assigned to $x_i$ is complemented, while others remain unchanged.

@@

{\bf Example 1}: Let $\ring$ be the set of rational numbers of the form $a\cdot 2^b$ where $a$ and $b$ are integers.
Let $\cost(x_i) = 1/2$ for all variables $x_i$.  The cost of every
assignment is then $1/2^{n}$, and the cost of a function is its
{\em density}, denoted $\density(f)$.  That is, the density of $f$ satisfies
$0 \leq \density(f) \leq 1$.  It
is the fraction of assignments for which the $f$
evaluates to 1, with $\density(\zero) = 0$ and $\density(\one) = 1$.  The density of a function
$f$ can be scaled by $2^n$ to compute the total number of models
$|\interpset{f}|$.  This is the core task of model counting.  Using
density as the metric, rather than the number of models, has the advantage that it does not vary when the
function is embedded in a larger domain $X' \supseteq X$.

{\bf Example 2}: Let $\ring$ be the set of rational numbers.  Assign a
{\em weight} $w_i$ for each variable $x_i$ such that $0 \leq w_i \leq
1$ and let $\cost(x_i) = w_i$.  This can support a restricted version of
{\em weighted} model counting.  The key restrictions are: 1) the
weight of an assignment equals the product of the weights of its
literals, and 2) the weight of a variable $x_i$ and its complement
$\obar{x}_i$ sum to 1.

@@

\begin{lemma}[Complementation]
\label{lemma:complementation}
  For separable cost function $\cost$ and Boolean function $f$:
  $\cost(\neg f) = \oneminus \cost(f)$.
\end{lemma}

\begin{lemma}[Variable-Partitioned Conjunction]
\label{lemma:conjunction}
  For separable cost function $\cost$ and Boolean functions $f_1$ and $f_2$ such that $\depend(f_1) \cap \depend(f_2) = \emptyset$:
    $\cost(f_1 \land f_2) = \cost(f_1) \cdot \cost(f_2)$.
\end{lemma}
We use the notation $f_1 \pand f_2$ to denote the conjunction of $f_1$ and
$f_2$ under the condition that $f_1$ and $f_2$ are defined over
disjoint sets of variables.

\begin{lemma}[Assignment-Partitioned Disjunction]
\label{lemma:disjunction}
  For separable cost function $\cost$ and Boolean functions $f_1$ and $f_2$ such that $\interpset{f_1} \cap \interpset{f_2} = \emptyset$:
    $\cost(f_1 \lor f_2) = \cost(f_1) + \cost(f_2)$.
\end{lemma}
We use the notation $f_1 \por f_2$ to denote the disjunction of $f_1$ and $f_2$ under the
condition that $f_1$ and $f_2$ hold for mutually exclusive assignments.

We will also find it useful to introduce a restricted version of the
if-then-else operator, denoted $\pite$.  That is, for functions $f_i$,
$f_t$, and $f_e$, we define $\pite(f_i, f_t, f_e) = (f_i \pand f_t)
\por (\neg f_i \pand f_e)$.  The two arguments to the disjunction are
guaranteed to satisfy the constraints for an assignment-partitioned
disjunction, since first argument can only be satisfied by assignments
for which $f_i$ evaluates to $\tautology$, while the second can only
be satisfied by assignments for which $f_i$ evaluates to $\nil$.  The
only restriction of the $\pite$ operator is that
$\depend(f_i) \cap \depend(f_t) = \emptyset$ and
$\depend(f_i) \cap \depend(f_e) = \emptyset$.

@@

\begin{table}
  \caption{Recursive Definition of Separable Schemas}
  \label{tab:schema}
A) Primitive Operations\\
\begin{center}
\begin{tabular}{cccc}
\toprule
  \makebox[15mm]{$S$}           & \makebox[30mm]{Restrictions} & \makebox[30mm]{$D(S)$} & \makebox[30mm]{$\interpset{S}$}\\
\midrule
  $\nil$  & None  & $\emptyset$    & $\emptyset$ \\
  $\tautology$        & None  & $\emptyset$    & $\uinterp$ \\
$x_i$           & None  & $\{ x_i \}$    & $\{ \interp | \interp(x_i) = \tautology \}$ \\
$\neg S_1$      & None  & $\depend(S_1)$ & $\uinterp - \interpset{S_1}$ \\
$S_1 \pand S_2$ & $\depend(S_1) \cap \depend(S_2) = \emptyset$ & $\depend(S_1) \cup \depend(S_2)$ & $\interpset{S_1} \cap \interpset{S_2}$ \\
$S_1 \por S_2$  & $\interpset{S_1} \cap \interpset{S_2} = \emptyset$ &  $\depend(S_1) \cup \depend(S_2)$ &  $\interpset{S_1} \cup \interpset{S_2}$ \\
\bottomrule
\end{tabular}
\end{center}
B) Derived Operations\\
\begin{center}
\begin{tabular}{cc}
\toprule
  $S$           & Definition \\
\midrule
\makebox[30mm]{$\pite(S_1, S_2, S_3)$} & \makebox[40mm]{$[S_1 \pand S_2] \por [\neg S_1 \pand S_2]$}\\
\bottomrule
\end{tabular}
\end{center}
\end{table}

A {\em separable schema} is a directed acyclic graph representing a
Boolean formula where the only allowed operations are $\neg$, $\pand$,
$\por$ and $\pite$.  We use a graph representation to allow sharing of
subformulas.  The set of schemas over a set of variables $\{x_1, x_2,
\ldots, x_n\}$ can be defined recursively, as is shown in Table
\ref{tab:schema}.  Each schema $S$ has an associated dependency set
$\depend(S)$ and an associated set of models $\interpset{S}$.  We
divide the operations into {\em basic} ones that are fundamental to
the representation, and {\em derived} ones that can be constructed
from basic operations.  The latter category consists of just the $\pite$ operation.

@@

Table~\ref{tab:normalize} shows a list of {\em normalizing}
transformations to simplify a separable schema.  These are mostly
straightforward---eliminating extra negations and removing constant
terms.  The most interesting are the two bottom rules, were we make
use of DeMorgan's Laws to simplify an $\pite$ operation when one of
the arguments is $\tautology$.  We will make use of this
transformation when using BDDs to convert a set of clauses into a
separable schema.

@@

The CRAT format draws its inspiration from the LRAT format for Boolean formulas and the
QRAT format for
quantified Boolean formulas (QBF).  The following are its key properties:
\begin{itemize}
\item As with LRAT, a clause can be added as long as either 1) it is
  blocked, or 2) it satisfies the RAT property with respect to a supplied sequence
  of earlier antecedent clauses.
\item  Extension variables can be introduced only according to the operations $\pand$, $\por$, and $\pite$.
\begin{itemize}
\item The checker tracks the dependency set for every input and
  extension variable.  When an extension variable is introduced based
  on the $\pand$ or $\pite$ operation, the dependency sets of its arguments must
  be disjoint.
\item When an extension variable is introduced based on the $\por$
  operation, the step must cite earlier steps providing a RAT proof
  that the two arguments are mutually exclusive.
\item Boolean complement is provided implicitly by allowing the
  arguments of the extension operations to be literals and not just
  variables.
\end{itemize}
\item A CRAT proof must show that the schema is logically equivalent
  to the input formula, not just that they are equisatisfiable.
  Therefore, each deletion step must also be show to be equivalence
  preserving, either because the clause is blocked or it follows from
  remaining clauses by the RAT property.
\item Unlike QRAT, it need not support universal quantification.
\end{itemize}

@@

Table~\ref{tab:crat:syntax} shows the set of proof rules for the CRAT
format.
As with other clausal proof formats, a variable is
represented by a positive integer $v$, with the first ones being input
variables and successive ones being extension variables.  Literal $l$
is represented by a signed integer, with $-v$ being the complement of
variable $v$.  Each clause is indicated by a positive integer
identifier $C$, with the first ones being input clauses and successive
ones being added clauses.  Clause identifiers must be totally ordered,
such that clause $C$ can only reference clauses $C'$ such that $C' <
C$.  However, clause identifiers need not be consecutive.

The first set of proof rules are similar to those in other clausal
proofs.  Our syntax optionally allows input clauses to be listed with
a rule of type {\tt i}.  Clauses can be added via blocked-clause and
RAT rules.  As described below, however, blocked clauses can only be added
to define $\pand$, $\por$, and $\pite$ operations.  The hints portion
of a blocked-clause addition lists all earlier clauses containing the
negated version of the pivot literal, with the clause IDs negated.
The hints portion of a RAT addition must contain a sequence of clause
IDs such that the added clause is RAT with respect to these clauses.
Clause deletion requires an explicit justification that the deleted
clauses is RAT with respect to other clauses.

@@

A CRAT proof follows the same general form as a QRAT dual proof---one
that ensures that each clause addition and each clause deletion preserves
equivalence.  Starting with the set of input clauses, it produces a
sequence of steps that both add and delete clauses.  Each addition
must be truth preserving and each deletion must be falsehood
preserving.  At the end, all input clauses must have been deleted, and
among the remaining clauses there must be only a single unit clause consisting of some variable or
its complement.  Except for trivial cases, the final literal will be
an extension variable or its complement.  That variable will indicate the root node of the schema.
Working from that root backward, the schema can be extracted from the CRAT file.

@@

As an illustration, consider the Boolean formula $x_1 \lor x_2 \lor
x_3$, represented by a single clause.  We cannot directly use the
$\por$ operation to form these disjunctions, since the sets of assignments
satisfying the individual literals are not disjoint.  Instead, we must
decompose this formula into a sequence of operations.  Figure~\ref{fig:c3:schema}A shows one such decomposition.
Circles in this schema
correspond to $\pand$ operations and squares to $\por$ operations.
The subscripts of the variables and the operator labels correspond to
the numbers of the input and extension variables in the CRAT proof.

@@

The logical representation can readily be converted into a schema for
computing the cost of the formula, given weight $w_i$ for each
variable $x_i$ for $1 \leq i \leq 3$.  This is illustrated in
Figure~\ref{fig:c3:schema}B\@.  This schema is valid for any cost
function, since it contains no abstraction operations.

Figure~\ref{fig:c3:crat} shows an annotated version of the CRAT proof
for this example.  Clause \#1 corresponds to the input formula, and
clauses \#2--\#13 are the defining clauses for the four operations.
Each of the two sum operations lists one of the earlier defining
clauses as a proof that its arguments are mutually exclusive.
Clause \#14 adds the unit clause corresponding to sum $s_7$,
indicating that the extension variable variable will evaluate to
$\tautology$  for any assigment that satisfies the input clause.  We
can write this as $C \turnstile s_7$ for input clause $C$.  The
deletion step at the end turns this around, showing that $s_7
\turnstile C_I$, and therefore the input clause can be deleted, This
completes a proof that extension variable $s_7$ is logically
equivalent to the input formula.

@@

\begin{figure}
  \centering{
    \begin{minipage}{0.4\textwidth}
      A) Schema   \\[2ex]
    \centering{\input{dd/p2}}
    \end{minipage}
    \begin{minipage}{0.4\textwidth}
    B) Cost Computation\\[2ex]
    \centering{\input{dd/p2-eval}}
    \end{minipage}
  }
\caption{Schema for Formula $\phi_2 = (x_1 \lor x_2 \lor x_3) \land (\obar{x}_1 \lor x_2)$ and its Cost Computation}
\label{fig:c3:schema}
\end{figure}

@@

As a more complex example, consider the Boolean formula given by the
conjunction of clauses $C_1 = x_1 \lor x_2 \lor x_3$ and $C_2 =
\obar{x}_1 \lor x_2$.  With this example, we also demonstrate the use
of DeMorgan's Laws to provide a more direct encoding of the problem.
The complement of the formula can be written in DNF as
$\obar{x}_1\,\obar{x}_2\,\obar{x}_3 \lor x_1\,\obar{x}_2$.  Each of
the two conjuncts can be formed using $\pand$ operations, and their
sum can be formed using a $\por$ operation.  In this example, we first
form a term $t_1$ that equals $C_1$ and a term $t_2$ that equals
$C_2$.  These are asserted as unit clauses in proof lines 9 and 13,
allowing the input clauses to be deleted.  Then we generate $t_3$ as
the conjunction of $t_1$ and $t_2$ and assert it as a unit clause.
Based on this, we can delete the unit clauses for terms $t_1$ and
$t_2$.  Term $t_3$ then becomes the root of the schematic representation.

@@

Given an arbitrary CNF formula, we can use BDD operations to generate
a schematic representation.  The proof generation can follow the
methods we have used for generating unsatisfiability proofs of Boolean
formulas~\cite{bryant:tacas:2021} and dual proofs of quantified
Boolean formulas~\cite{bryant:cade:2021}.  The key idea 
is to use extended resolution to encode the semantics of the BDD
nodes as part of the proof.  Here we can further decompose each $\ite$
(short for ``if-then-else'') operation representing a BDD node into
two $\pand$ and one $\por$ operation.  That is, we can encode
$\ite(x,A,B)$ as $(x \pand A) \por (\obar{x} \pand B)$, where $x$ is an
input variable and $A$ and $B$ are subformulas.  The sum operation trivially satisfies the disjoint assignment requirement, since $x$ has positive polarity in the first product and negative polarity in the second.
We must also
make sure that $x$ is not in the dependency set of either $A$ or $B$.
For ordered BDDs,~\cite{Bryant:1986} this property holds, because the
variables in $A$ and $B$ will be greater in the variable ordering than
$x$.

The BDD representations of many of the formulas occuring in
model-counting problems are far to large for this approach to be
practical.  One refinement of the approach, implemented by the ADDMC
model counter, is to abstract subformulas, keeping track only of the
number of models they can have and representing a number of subformulas
with a single ADD leaf node.  Our abstraction operation is intended to
support this capability.  We can
introduce a new variable to represent the cost function for some
subformula.  If subsequent formulas are guaranteed to yield the same
cost, then these can alias to the earlier variable.  However, it is
not clear how to prove that this aliasing preserves equivalence.

@@

\begin{itemize}
\item Proof Framework
  \begin{itemize}
  \item Generalities and details of the format
  \item What should be the final state?
  \item When can blocked clauses be deleted?
  \item How can abstraction be incorporated?
  \end{itemize}

@@

\item CDCL-based
  \begin{itemize}
  \item Prototype
  \item C/C++
  \end{itemize}
\end{itemize}

\end{itemize}

>>>>

A {\em Boolean function} $f:2^X \rightarrow \{0,1\}$ can be
characterized by the set of assignments for which the function
evaluates to 1: $\interpset{f} = \{ \interp | f(\interp) = 1\}$.  Let
$\one$ denote the Boolean function that assigns value 1 to every
assignment, and $\zero$ denote the assignment that assigns value 0 to
every assignment.  These are characterized by the universal assignment set $\uinterp$ and the empty
assignment set $\emptyset$, respectively.

From
this we can define the {\em negation} of function $f$ as the function
$\boolnot f$ such that
$\interpset{\boolnot f} = \uinterp - \interpset{f}$.
We can also define the conjunction and disjunction operations over functions $f_1$ and $f_2$ as characterized by the sets
$\interpset{f_1 \booland f_2} = \interpset{f_1} \cap \interpset{f_2}$ and
$\interpset{f_1 \boolor f_2} = \interpset{f_1} \cup \interpset{f_2}$.

For assignment $\interp$ and a Boolean formula $\phi$ over $X$, we
use the notation $\interp\subs{x_i}{\phi}$ to denote the assignment
$\interp'$, such that $\interp'(x_j) = \interp(x_j)$ for all $j \not = i$
and $\interp'(x_i) = \interp(\phi)$, where $\interp(\phi)$ is the value obtained by evaluating formula $E$ with each variable assigned the value given by $\interp$.
In particular, the notation
$\interp(\subs{x_i}{\obar{x}_i})$
indicates the assignment in which the value
assigned to $x_i$ is negated, while others remain unchanged.

@@

{\bf Example 1}: Let $\ring$ be the set of rational numbers of the form $a\cdot 2^b$ where $a$ and $b$ are integers.
Let $\cost(x_i) = 1/2$ for all variables $x_i$.  The cost of every
assignment is then $1/2^{n}$, and the cost of a function is its
{\em density}, denoted $\density(f)$.  That is, the density of $f$, for which
$0 \leq \density(f) \leq 1$, 
is the fraction of assignments for which $f$
evaluates to 1, with $\density(\zero) = 0$ and $\density(\one) = 1$.  The density of a function
$f$ can be scaled by $2^n$ to compute the total number of models
$|\interpset{f}|$.  This is the core task of model counting.  Using
density as the metric, rather than the number of models, has the advantage that it does not vary when the
function is embedded in a larger domain $X' \supseteq X$.

{\bf Example 2}: Let $\ring$ be the set of rational numbers.  Assign a
{\em weight} $w_i$ to each variable $x_i$ such that $0 \leq w_i \leq
1$ and let $\cost(x_i) = w_i$.  This implements {\em weighted model counting}
under the restrictions that: 
1) the
weight of an assignment equals the product of the weights of its
literals, and 2) the weight of a variable $x_i$ and its complement
$\obar{x}_i$ sum to 1.

@@

\begin{proposition}[Complementation]
\label{prop:complementation}
  For separable cost function $\cost$ and Boolean function $f$:
  $\cost(\neg f) = 1 - \cost(f) = \oneminus \cost(f)$.
\end{proposition}

\begin{proposition}[Variable-Partitioned Conjunction]
\label{prop:conjunction}
  For separable cost function $\cost$ and Boolean functions $f_1$ and $f_2$ such that $\depend(f_1) \cap \depend(f_2) = \emptyset$:
    $\cost(f_1 \land f_2) = \cost(f_1) \cdot \cost(f_2)$.
\end{proposition}
We use the notation $f_1 \pand f_2$ to denote the conjunction of $f_1$ and
$f_2$ under the condition that $f_1$ and $f_2$ are defined over
disjoint sets of variables.

\begin{proposition}[Assignment-Partitioned Disjunction]
\label{prop:disjunction}
  For separable cost function $\cost$ and Boolean functions $f_1$ and $f_2$ such that $\interpset{f_1} \cap \interpset{f_2} = \emptyset$:
    $\cost(f_1 \lor f_2) = \cost(f_1) + \cost(f_2)$.
\end{proposition}
We use the notation $f_1 \por f_2$ to denote the disjunction of $f_1$ and $f_2$ under the
condition that $f_1$ and $f_2$ hold for mutually exclusive assignments.

We also find it useful to introduce a restricted version of the
if-then-else operator, denoted $\pite$.  For functions $f_1$,
$f_2$, and $f_3$, we define $\pite(f_1, f_2, f_3) = (f_1 \pand f_2)
\por (\neg f_1 \pand f_3)$.  The two arguments to the disjunction are
guaranteed to satisfy the requirements for an assignment-partitioned
disjunction, since the first argument can only be satisfied by assignments
for which $f_1$ evaluates to $\tautology$, while the second can only
be satisfied by assignments for which $f_1$ evaluates to $\nil$.  The
only restriction of the $\pite$ operator is that
$\depend(f_1) \cap \depend(f_2) = \emptyset$ and
$\depend(f_1) \cap \depend(f_3) = \emptyset$.

@@

\begin{table}
  \caption{Recursive Definition of Separable Schemas}
  \label{tab:schema}
A) Basic Operations\\
\begin{center}
\begin{tabular}{cccc}
\toprule
  \makebox[15mm]{$S$}           & \makebox[30mm]{Restrictions} & \makebox[30mm]{$D(S)$} & \makebox[30mm]{$\interpset{S}$}\\
\midrule
  $\nil$  & None  & $\emptyset$    & $\emptyset$ \\
  $\tautology$        & None  & $\emptyset$    & $\uinterp$ \\
$x_i$           & None  & $\{ x_i \}$    & $\{ \interp | \interp(x_i) = \tautology \}$ \\
$\neg S_1$      & None  & $\depend(S_1)$ & $\uinterp - \interpset{S_1}$ \\
$S_1 \pand S_2$ & $\depend(S_1) \cap \depend(S_2) = \emptyset$ & $\depend(S_1) \cup \depend(S_2)$ & $\interpset{S_1} \cap \interpset{S_2}$ \\
$S_1 \por S_2$  & $\interpset{S_1} \cap \interpset{S_2} = \emptyset$ &  $\depend(S_1) \cup \depend(S_2)$ &  $\interpset{S_1} \cup \interpset{S_2}$ \\
\bottomrule
\end{tabular}
\end{center}
B) Composite Operations\\
\begin{center}
\begin{tabular}{cc}
\toprule
  $S$           & Definition \\
\midrule
\makebox[30mm]{$\pite(S_1, S_2, S_3)$} & \makebox[40mm]{$[S_1 \pand S_2] \por [\neg S_1 \pand S_2]$}\\
\bottomrule
\end{tabular}
\end{center}
\end{table}

A {\em separable schema} is a directed acyclic graph representing a
Boolean formula where the only allowed operations are $\neg$, $\pand$,
$\por$ and $\pite$.  We use a graph representation to allow sharing of
subformulas.  The set of schemas over a set of variables
$\{x_1, x_2, \ldots, x_n\}$
can be defined recursively, as is shown in Table
\ref{tab:schema}.  Each schema $S$ has an associated dependency set
$\depend(S)$ and an associated set of models $\interpset{S}$.  We
divide the operations into {\em basic} ones that are fundamental to
the representation, and {\em composite} ones that can be constructed
from basic operations.  The latter category consists of just the $\pite$ operation.

@@

Table~\ref{tab:normalize} shows a list of {\em normalizing}
transformations to simplify a separable schema.  These are mostly
straightforward---eliminating extra negations and removing constant
terms.  The most interesting are the two bottom rules, which
use DeMorgan's Laws to simplify an $\pite$ operation when one of
the arguments is $\tautology$ into a $\pand$ operation.  We will make use of this
transformation when using BDDs to convert a set of clauses into a
separable schema.

@@

The CRAT format draws its inspiration from the LRAT format for Boolean formulas and the
QRAT format for
quantified Boolean formulas (QBF).  The following are its key properties:
\begin{itemize}
\item As with LRAT, a clause can be added as long as either 1) it is
  blocked, or 2) it satisfies the RAT property with respect to a supplied sequence
  of earlier antecedent clauses.  However, blocked clauses can only be added when defining a schema operation.
\item  Extension variables can be introduced only according to the operations $\pand$, $\por$, and $\pite$.
\begin{itemize}
\item The checker tracks the dependency set for every input and
  extension variable.  When an extension variable is introduced based
  on the $\pand$ or $\pite$ operation, the dependency sets of its arguments must
  be disjoint.
\item When an extension variable is introduced based on the $\por$
  operation, the step must cite earlier steps providing a RAT proof
  that the two arguments are mutually exclusive.
\item Boolean complement is provided implicitly by allowing the
  arguments of the extension operations to be literals and not just
  variables.
\end{itemize}
\item A CRAT proof must show that the schema is logically equivalent
  to the input formula, not just that they are equisatisfiable.
  Therefore, each deletion step must also be shown to be equivalence
  preserving, either because the clause is blocked or it follows from
  remaining clauses by the RAT property.
\item Unlike QRAT, it need not support universal quantification.
\end{itemize}

@@

Table~\ref{tab:crat:syntax} shows the set of proof rules for the CRAT
format.
As with other clausal proof formats, a variable is
represented by a positive integer $v$, with the first ones being input
variables and successive ones being extension variables.  Literal $l$
is represented by a signed integer, with $-v$ being the complement of
variable $v$.  Each clause is indicated by a positive integer
identifier $C$, with the first ones being the IDs of the input clauses and successive
ones being the IDs of added clauses.  Clause identifiers must be totally ordered,
such that clause $C$ can only reference clauses $C'$ such that $C' <
C$.  Clause identifiers need not be consecutive.

The first set of proof rules are similar to those in other clausal
proofs.  Our syntax optionally allows input clauses to be listed with
a rule of type {\tt i}.  Clauses can be added via a blocked-clause addition (command {\tt ab})
or a RAT addition (command {\tt dr}).
As described below, however, blocked clauses can only be added
to define $\pand$, $\por$, and $\pite$ operations.  The hints portion
of a blocked-clause addition lists all earlier clauses containing the
negated version of the pivot literal, with the clause IDs negated.
The hints portion of a RAT addition must contain a sequence of clause
IDs such that the added clause is RAT with respect to these clauses.
Similarly, a clause can be deleted if it blocked with respect to one
of its literals (command {\tt db}), or because it is RAT (command {\tt
  dr}) The hints portion of a RAT deletion operation must be an
ordered list of clause IDs justifying the deletion.

@@

A CRAT proof follows the same general form as a QRAT dual proof---one
that ensures that each clause addition and each clause deletion
preserves equivalence.  Starting with the set of input clauses, it
produces a sequence of steps that both add and delete clauses.  Each
addition must be truth preserving and each deletion must be falsehood
preserving.  At the end, all input clauses must have been deleted, and
among the remaining clauses there must be only a single unit clause
consisting of some variable or its complement.  Except for trivial
cases, the final literal will be an extension variable or its
complement.  That literal will indicate the root of the schema as the
(possibly negated) outpuf of a schema operation.  Working from that
root backward, the schema can be extracted from the CRAT file.

@@

As an illustration, consider the Boolean formula
$\phi_1 = x_1 \lor x_2 \lor x_3$,
represented by a single clause.  We cannot directly use the
$\por$ operation to form these disjunctions, since the sets of assignments
satisfying the individual literals are not disjoint.  Instead, we must
decompose this formula into a sequence of operations.
Figure~\ref{fig:c3:schema}A shows one such decomposition.
The subscripts of the variables and the operator labels correspond to
the numbers of the input and extension variables in the CRAT proof.
Edges marked with dots indicate Boolean negation.

@@

The logical representation can readily be converted into a formula for
computing $\cost(\phi_1)$, the cost of formula $\phi_1$, given a weight
$w_i$ for each variable $x_i$ for $1 \leq i \leq 3$.  This is
illustrated in Figure~\ref{fig:c3:schema}B\@.  Note how the Boolean
negations become $\oneminus$ operations.  This formula is valid for
any cost function.

Figure~\ref{fig:c3:crat} shows an annotated version of the CRAT proof
for this example.  Clause \#1
is the input clause, and
clauses \#2--\#13 are the defining clauses for the four operations.
Each of the two sum operations lists one of the earlier defining
clauses as a proof that its arguments are mutually exclusive.
Proof clause \#14 adds the unit clause for the extension variable $s_7$.
We refer to the literal representing formula $\phi_1$ as $t_1$,
and we therefore have $t_1 = s_7$.
The unit clause indicates that extension variable $s_7$ will evaluate to
$\tautology$  for any assigment that satisfies the formula.  We
can write this as $\phi_1 \turnstile t_1$.  The
deletion step at the end turns this around, showing that
$t_1 \turnstile \phi_1$,
and therefore the input clause can be deleted, This
completes a proof that $t_1$ is logically
equivalent to the input formula.

@@

\begin{figure}
  \centering{
    \begin{minipage}{0.4\textwidth}
      A) Schema   \\[2ex]
    \centering{\input{dd/p2}}
    \end{minipage}
    \begin{minipage}{0.4\textwidth}
    B) Cost Computation\\[2ex]
    \centering{\input{dd/p2-eval}}
    \end{minipage}
  }
\caption{Schema for Formula $\phi_2 = (x_1 \lor x_2 \lor x_3) \land (\obar{x}_1 \lor x_2)$ and its Cost Computation}
\label{fig:p2:schema}
\end{figure}

@@

As a more complex example, consider the Boolean formula given by the
conjunction of clauses $C_1 = x_1 \lor x_2 \lor x_3$ and $C_2 =
\obar{x}_1 \lor x_2$.  With this example, we also demonstrate the use
of DeMorgan's Laws to provide a more compact encoding of the formula.
The complement of the formula can be written in DNF as
$\obar{x}_1\,\obar{x}_2\,\obar{x}_3 \lor x_1\,\obar{x}_2$.  Each of
the two conjuncts can be formed using $\pand$ operations, and their
sum can be formed using a $\por$ operation.  In this example, we first
form a term $t_1$ that equals $C_1$ and a term $t_2$ that equals
$C_2$.  These are asserted as unit clauses in proof lines 9 and 13,
allowing the input clauses to be deleted.  Then we generate $t_3$ as
the conjunction of $t_1$ and $t_2$ and assert it as a unit clause.
Based on this, we can delete the unit clauses for terms $t_1$ and
$t_2$.  Term $t_3$ then becomes the root of the schematic representation.

@@

Given an arbitrary CNF formula, we can use BDD operations to generate
a schematic representation.  The proof generation can follow the
methods we have used for generating unsatisfiability proofs of Boolean
formulas~\cite{bryant:tacas:2021} and dual proofs of quantified
Boolean formulas~\cite{bryant:cade:2021}.  The key idea is to use
extended resolution to encode the semantics of the BDD nodes as part
of the proof.  Each BDD node defines an ITE operation.  As the
transformations of Table~\ref{tab:normalize} indicate, each of these
will be represented in the proof by a literal, an $\pand$ operation,
or an $\pite$ operation.  The ITE operation for a node has an input
variable $x_i$ as its first argument.  In an ordered BDD, the
dependency sets of the two children will include only variables
greater than $x_i$ in the ordering.  Therefore, these operations will
satisfy the partitioned-variable requirements of the  $\pand$ and $\pite$ operations.

@@

\begin{itemize}
\item Proof Framework
  \begin{itemize}
  \item Generalities and details of the format
  \item Can some form of abstraction be incorporated?
    \begin{itemize}
    \item Want to abstract a subformula to consider only on its cost and dependency set
    \item Could represent with fresh extension variable
    \item But how to prove logical equivalence?
    \end{itemize}
  \end{itemize}

@@

  \item SDD-based
  \begin{itemize}
  \item Bottom-up
  \item Top-down
  \end{itemize}

\item Others?
\end{itemize}
\end{itemize}
